---
title: "Second Data Product"
output: html_notebook
---

The second data product for my university course.
It's aim is to predict housing prices using regression.  
The used data is the Ames Housing dataset, from kaggle.com
It's split into two files, a train.csv and a test.csv.  
Created by: Dobosi Péter MW79ON



## Business Understanding:

First let's decide what the problem exactly is, and what do we want to achieve.
The exact overview of the problem can be read here: 
https://www.kaggle.com/c/house-prices-advanced-regression-techniques/overview

In summary, we have to create a model, which can predict the sale price of a
house, given it's parameters, using regression.

My plan is the following:

1. Explore the data.
2. Clean and transform it into a usable format.
3. Choose, build and verify a model.
4. Evaluate it.
5. Use it to predict on the final dataset.


First things first, let's load the data and take a quick look at it:
```{r}
data <- read.csv("./train.csv")
data
```

We have more than 1400 objects and 79 parameters. We have NAs and everything
is simply either characters or integers.  
To properly prepare the data we have to 
handle the empty values one way or another, and also convert the characters to
factors.

It seems like we have a couple of parameters that are mostly NAs:
```{r}
alley_na_percentage <- sum(is.na(data$Alley))/1460*100
pool_na_percentage <- sum(is.na(data$PoolQC))/1460*100
fence_na_percentage <- sum(is.na(data$Fence))/1460*100
feature_na_percentage <- sum(is.na(data$MiscFeature))/1460*100

paste("Percentage of NAs in the field Alley:", alley_na_percentage, "%")
paste("Percentage of NAs in the field PoolQC:", pool_na_percentage, "%")
paste("Percentage of NAs in the field Fence:", fence_na_percentage, "%")
paste("Percentage of NAs in the field MiscFeature:", feature_na_percentage, "%")
```

With the help of data_description.txt we can decipher what do these mean:  
(Only the relevant parts here, read the rest from the file if you are interested.)

Alley:  
Type of alley access to property.

       Grvl	Gravel
       Pave	Paved
       NA 	No alley access

PoolQC:  
Pool quality.

       Ex	Excellent
       Gd	Good
       TA	Average/Typical
       Fa	Fair
       NA	No Pool

Fence:  
Fence quality.
		
       GdPrv	Good Privacy
       MnPrv	Minimum Privacy
       GdWo	Good Wood
       MnWw	Minimum Wood/Wire
       NA	No Fence

MiscFeature:  
Miscellaneous feature not covered in other categories.
		
       Elev	Elevator
       Gar2	2nd Garage (if not described in garage section)
       Othr	Other
       Shed	Shed (over 100 SF)
       TenC	Tennis Court
       NA	None


As we can see, they don't mean that we have no information on those parameters
of the buildings. Rather their meaning is simply that they lack the things
described by those parameters. This is important information, we can't just drop, or guess them from based on the others.



## Exploratory Data Analysis:

Let's find out more about the data's characteristics!  
Let's take a quick look at the parameters of the dataset:
```{r}
names(data)
```
As we can see, we have roughly 80 parameters.  
Let's take a look at the types of the parameters:
```{r}
str(data)
```
All of our parameters are either character or integer vectors.  
Let's take a look at the summary:
```{r}
summary(data)
```
A lot of our attributes are character vectors, which we can't summarize this way.


## One dimensional examination

In which we find out more about given parameters of the data.  
Let's take a visual look:
```{r}
plot(data$LotArea)
plot(data$LotArea, ylim = c(1000, 20000))
```

We can see, that in terms of Area, most of the properties are between 1000 and
20000 square feet, with a few outliers.  
Let's check out how many houses were built in each year:

```{r}
hist(data$YearBuilt)
```

We can also see, a tendency towards newly built homes.

```{r}
plot(table(data$Fireplaces))
grid()
```
Multidimensional examination:

Now let's take a look at multiple parameters at the same time:
```{r}
library(car)
scatterplot(data$YearBuilt, data$SalePrice, regLine = list(col="green"), smooth=list(col.smooth="red", col.spread="black"))
```

We might see some kind of exponential pattern, given the higher prices of newly built homes.

```{r}
scatterplot(data$LotArea, data$SalePrice, xlim = c(1000, 20000), regLine = list(col="green"), smooth=list(col.smooth="red", col.spread="black"))
```

We can't see any clear correlation between the area and the price of a property.


Just out of curiosity, I tried to draw the feature plot of the data.
To my surprise, with a couple of tweaks, and given some time it actually worked:
```{r}

# This took a couple minutes, but it worked. It's about 64k square meters.
# The plot is basically unreadable, but it shows that there is correlation
# between a couple of the parameters.

# pdf(file = "/home/peter/test.pdf",
#     width = 10000,
#     height = 10000)

# plot(data)

# dev.off()
```

Although we couldn't really learn anything from it, due to it's size it's 
unreadable.


Let's check out the covariance matrix:
```{r}
# cov(data)
```

This doesn't work because there are non numeric or logical values in our
dataframe still. It's time we cleaned up the data a bit. But before we do that,
let's try to take a look at the correlations between a couple other parameters.

```{r}
library(corrplot)
cors <- cor(data[,c(5,19,20,39,47,63,72,77,81)], use = "complete.obs")
corrplot(cors, type = "lower")
```

We can see correlation between the price and a couple of parameters, such as the
size of the living area.




## Data Cleaning:

There is a lot to do. We have NAs, and non numeric values everywhere.
Let's start by dealing with the NAs first.  
We need to find out which columns contain any NAs:
```{r}
na_cols <- names(which(colSums(is.na(data)) > 0))
na_cols
```
We have 19 columns containing NAs, let's find out more about them.  
Let's find out how many NAs do these columns have:
```{r}
get_na_count <- function(column_name) {
    sum(is.na(data[column_name]))
}

na_counts <- data.frame(sapply(na_cols, get_na_count))

library(data.table)
na_stats <- transpose(na_counts)

colnames(na_stats) <- na_cols
rownames(na_stats) <- c("NA count")

calc_na_percentage <- function(column_name) {
    get_na_count(column_name = column_name)/nrow(data) * 100
}

na_stats[nrow(na_stats) + 1,] = sapply(na_cols, calc_na_percentage)
rownames(na_stats) <- c("NA count", "NA percentage")

na_stats
```

As we can see, we have some parameters that are mostly NAs, while others only 
contain a few of them.  
Let's deal with them appropriately, now that we know more about them.  
First, the LotFrontage parameter:
```{r}
unique(data$LotFrontage)
```


The description doesn't say anything about NAs in this parameter, but as we can
see, there aren't any zeros here. So I'll assume that NAs mean zero here as well,
as it does in most of the other parameters.  
Let's fill them in now:
```{r}
data[is.na(data$LotFrontage),]$LotFrontage <- 0
```


The Alley parameter:  
The data_description.txt says that NAs in this parameter mean, that there is no
alley access, to the given property.
```{r}
unique(data$Alley)
```

Later I'll probably convert all character
vectors to factors, so let's leave this as is.  
Now for the Masonry veneer type:
```{r}
ms_types <- unique(data$MasVnrType)
ms_types
```

We have a handful of NAs, but here they do not simply mean that there is no such
thing as what's being described by the parameter. We have to actually fill them
in.  
Let's do so by the most frequent value:

```{r}
get_ms_count <- function(unique_value){
    sum(data$MasVnrType == unique_value, na.rm = T)
}

sapply(ms_types, get_ms_count)
```

As we can see, the most common option is None, so let's assume that NAs are None:
```{r}
data[is.na(data$MasVnrType),]$MasVnrType <- "None"
```

We have to do the same for Masonry veneer area as well, but with 0s this time:
```{r}
data[is.na(data$MasVnrArea),]$MasVnrArea <- 0
```

BsmtQual is next: 
```{r}
unique(data$BsmtQual)
```

According to the description, NAs here mean, that the property has no basement.
Let's leave this as is.  
The same is true for BsmtCond, BsmtExposure, BsmtFinSF1, BsmtFinType1 and BsmtFinType2.  
Electrical is up next:
```{r}
elec_types <- unique(data$Electrical)
elec_types
```

The description doesn't say anything about the one missing value, so let's fill
it with the most frequent value:
```{r}

# TODO I need to change these to reusable methods.

get_elec_count <- function(unique_value){
    sum(data$Electrical == unique_value, na.rm = T)
}

sapply(elec_types, get_elec_count)
```

As we can see, the Standard Breaker is the most common, let's assume the missing
value is that:
```{r}
data[is.na(data$Electrical),]$Electrical <- "SBrkr"
```

FireplaceQu is next:  
The description says that NAs here mean that there is no fireplace, so let's 
leave this as is.  
The same deal for all the parameters describing the garages.  
PoolQC and Fence also behave the exact same way.  
Finally the last one, MiscFeature. This one is similar, NAs simply mean that
there aren't any misc features.  

Finally after all this hard work, we shouldn't have any NAs left in our
dataframe, where they don't make any sense
Let's check whether that's true:
```{r}
names(which(colSums(is.na(data)) > 0))
```

It is!

After we've dealt with all of the NAs, let's check whether everything is the 
correct type:
```{r}
str(data)
```

Nothing seems out of order, but we still have a bunch of character vectors.
We need to encode them in a way, that our models can use. Let's convert them to
factors. This way R can automatically dummy code them when building models.

First things first, we have to find out which parameters are strings, so we can
know which ones to convert to factors:
```{r}
char_parms <- colnames(data[sapply(data, is.character)])
char_parms
```

As we can see, we have a bit more than 40 parameters which are characters.
Let's convert them to factors:
```{r}
data[char_parms] <- lapply(data[char_parms], factor)
```

Let's check whether we were successful:
```{r}
str(data)
```

We were!  
  
Now, after cleaning the data, let's check out the correlations, to see, which
parameters should we pay more attention to.  
First let's see for the numeric values:
```{r}
num_parms <- colnames(data[sapply(data, is.numeric)])
num_parms

numcors <- cor(data[,num_parms], use = "complete.obs")
corrplot(numcors, type = "lower")
```
This is hard to read, but we can already see that we don't need all of these 
parameters.  
Let's check out the more relevant ones:
```{r}
relevant_names <- names(numcors[38,numcors[38,] > 0.5])

relcors <- cor(data[,relevant_names], use = "complete.obs")
corrplot(relcors, type = "lower")
```
Let's check them out:
```{r}
scatterplot(data$OverallQual, data$SalePrice, regLine = list(col="green"), smooth=list(col.smooth="red", col.spread="black"))

scatterplot(data$GrLivArea, data$SalePrice, xlim = c(250, 3000), ylim = c(0, 500000), regLine = list(col="green"), smooth=list(col.smooth="red", col.spread="black"))

scatterplot(data$GrLivArea, data$TotRmsAbvGrd, regLine = list(col="green"), smooth=list(col.smooth="red", col.spread="black"))

scatterplot(data$OverallQual, data$GrLivArea, regLine = list(col="green"), smooth=list(col.smooth="red", col.spread="black"))

pairs(data[,c("SalePrice", "GrLivArea", "TotRmsAbvGrd")])
```

We can see a couple of obvious correlations, that don't mean anything, such as:
between the GarageArea and GarageCars, and GrLivArea and TotRmsAbvGrd.
  
Let's calculate a new parameter, the price per square feet:
```{r}
data$ppsqf <- data$SalePrice / data$GrLivArea
scatterplot(data$OverallQual, data$ppsqf, regLine = list(col="green"), smooth=list(col.smooth="red", col.spread="black"))
```
Let's check out the correlation between this new parameters and the old ones:
```{r}
relevant_names2 <- c(relevant_names, "ppsqf")
relevant_names2

relcors2 <- cor(data[,relevant_names2], use = "complete.obs")
corrplot(relcors2, type = "lower")
```
It seems as the price per square feet has risen over the years. Let's find it out:
```{r}
scatterplot(data$YearBuilt, data$ppsqf, regLine = list(col="green"), smooth=list(col.smooth="red", col.spread="black"))
```
We were right.


## Model Building:

Let's build some models. I would like to use the caret package to build a linear
and an exponential regression model.  
Let's create the data partitions first:
```{r}
library(caret)

target <- data$SalePrice
trainIdx <- createDataPartition(target, p = .75)
traindata <- data[trainIdx$Resample1,]
testdata <- data[-trainIdx$Resample1,]
```

```{r}
str(traindata)
```

After creating the partitions, let's build the model.  
First let's just use one parameter:
```{r}
model <- lm(SalePrice~OverallQual, data = traindata)
summary(model)
```



```{r}
plot(model)
shapiro.test(model$residuals)
confint(model)
cor(traindata$SalePrice, model$fitted.values)
model
```
```{r}
prediction <- predict(model, testdata, type="response")
model_output <- cbind(testdata, prediction)

model_output$log_prediction <- log(model_output$prediction)
model_output$log_SalePrice <- log(model_output$SalePrice)

rmse <- function(fittedvals, truevals){
  sqrt(mean((fittedvals - truevals)^2))
}

rmse(model_output$log_SalePrice,model_output$log_prediction)
```
As we can see, our model isn't any good. Let's try a different approach, with 
more parameters:
```{r}
model2 <- lm(SalePrice~OverallQual+GrLivArea, data = traindata)
summary(model2)
```
Let's evaluate it:
```{r}
plot(model2)
shapiro.test(model2$residuals)
confint(model2)
cor(traindata$SalePrice, model2$fitted.values)
model2
```
```{r}
prediction2 <- predict(model2, testdata, type="response")
model2_output <- cbind(testdata, prediction2)

model2_output$log_prediction <- log(model2_output$prediction)
model2_output$log_SalePrice <- log(model2_output$SalePrice)

rmse(model2_output$log_SalePrice, model2_output$log_prediction)
```

This somehow actually worsened our model, I'm not exactly sure why.
Anyway, let's try to give it more parameters:

```{r}
model3 <- lm(SalePrice~OverallQual+YearBuilt+YearRemodAdd+TotalBsmtSF+X1stFlrSF+
               GrLivArea+FullBath+TotRmsAbvGrd+GarageCars+GarageArea+ppsqf, data = traindata)
summary(model3)
```
Linear regression models are clearly not the way to go. Let's try some exponential
ones.
```{r}
model4 <- lm(log(SalePrice) ~ OverallQual, data = traindata)
summary(model4)
```

```{r}
plot(model4)
shapiro.test(model4$residuals)
confint(model4)
cor(traindata$SalePrice, model4$fitted.values)
model4
```
As we can see, this model is fairly better than our previous attempts.
Let's try the same thing, with more parameters.

```{r}
model5 <- lm(log(SalePrice)~OverallQual+YearBuilt+YearRemodAdd+TotalBsmtSF+X1stFlrSF+
               GrLivArea+FullBath+TotRmsAbvGrd+GarageCars+GarageArea+ppsqf, data = traindata)
summary(model5)
```

```{r}
plot(model5)
shapiro.test(model5$residuals)
confint(model5)
cor(traindata$SalePrice, model5$fitted.values)
model5
```

As we can see, our model did not improve, on the opposite, it worsened.

I would love to continue working on this exercise, improving my models, examining
the target parameter in relation to groups of objects, such as city areas, or 
types of buildings, but sadly I'm out of time.

As a last thing, I tried to check out the whether the prices and the sizes of 
living areas are affected by the neighborhood. We can only see some correlation
in the outliers.

```{r}
plot(data$GrLivArea, data$SalePrice, col=data$Neighborhood)
```

As a conclusion, I still have a lot to learn and would require a lot more time to
properly solve the problem. My current best model only uses a single parameter.
If I have the time for it in the future, I'll return to try to solve it properly.



